---
title: "A Descriptive Analysis of County-level Confirmed COVID-19 Cases in U.S."
author: "NYU A3SR"
output: 
  html_document:
    df_print: paged
    theme: paper
    highlight: kate
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    number_sections: false
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
if(!requireNamespace("dplyr"))
  install.packages("dplyr", repos = "https://cloud.r-project.org")
if(!requireNamespace("data.table"))
  install.packages("data.table", repos = "https://cloud.r-project.org")
if(!requireNamespace("here"))
  install.packages("here", repos = "https://cloud.r-project.org")
if(!requireNamespace("ggplot2"))
  install.packages("ggplot2", repos = "https://cloud.r-project.org")
require(data.table)
require(dplyr)
library(here)
require(ggplot2)
```

## Introduction

This descriptive analysis illustrates how county-level confirmed [COVID-19](https://www.who.int/emergencies/diseases/novel-coronavirus-2019) cases were developed in the United States, starting from January, 2020. The data we used in this study come from [USA FACTS](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/). The data files are updated on a daily basis and are stored in time stamp formats. We applied multiple statistical tools to analyze and visualize the outbreak development. 

## Creating Data Files

Through data processing, we created two data files for the analysis:

1. Daily nation-wide total confirmed cases

2. Most recent confirmed cases by county (as of 03-28-2020)

```{r loading-data-hide, include=FALSE}
# Data Name: covid_confirmed_usafacts_03282020.csv
# Path: data/raw/case data/usa-facts/03-28-2020/

# Load -------------------------------------------------------------------------
here::here()
dat <- fread("../data/raw/case data/usa-facts/03-28-2020/covid_confirmed_usafacts_03282020.csv")
```

## Descriptive Analysis

### Data Structure {.tabset}

Data are ordered in time series. Starting from Jan. 22, 2020 (1/22/2020), county-level confirmed cases were recorded daily. Each value time stamp variables is cumulative. For example, the value of confirmed cases in Queens, NY on Mar. 28, 2020 is calculated by adding cumulative confirmed cases on Mar. 27, 2020 with new cases confirmed on Mar. 28. Therefore, to get daily new cases, we can simply subtract the value on a particular day with the value in the previous day. 

#### Overview

Top 5 rows of raw data

```{r data-overview-1, warning=FALSE, message=FALSE}
head(dat, 5)
```

Bottom 5 rows of raw data

```{r data-overview-2, warning=FALSE, message=FALSE}
tail(dat, 5)
```

```{r data-overview-hide, include=FALSE}
# Check if there are missing values
table(is.na(dat)) 
# Review the distribution of confirmed cases as of 03-28-2020
summary(dat[, `3/28/2020`]) 

# Queens County, NY has the highest total as of 03-28-2020, 
# with 9,831 confirmed cases.
dat$`County Name`[dat$`3/28/2020` == max(dat$`3/28/2020`)]
dat$State[dat$`3/28/2020` == max(dat$`3/28/2020`)]
max(dat$`3/28/2020`)

# 1,325 counties have zero confirmed case as of 03-28-2020
table(dat$`3/28/2020` == 0)
```

#### Daily Nation-wide Total

A summary table of nation-wide daily total confirmed cases is attached. 

```{r data-daily-total, warning=FALSE, message=FALSE}
# Create a table showing daily total confirmed cases, nation-wide
confirmed_cases_per_day <- dat[, 5:71]
date <- as.Date(paste("0", colnames(confirmed_cases_per_day) , sep = ""), "%m/%d/%y")

total_confirmed_daily <- lapply(dat[, 5:71], FUN = sum)
total_confirmed_daily <- t(as.matrix(as.data.frame(total_confirmed_daily)))
total_confirmed_daily <- data.frame(date = date, 
                                    total_confirmed = total_confirmed_daily)
row.names(total_confirmed_daily) <- NULL

total_confirmed_daily
```

#### Most Recent by County

```{r data-show-3, warning=FALSE, message=FALSE}
# Create a tidy datafile to include only 
# county, state, and most recent confirmed cases.
total_cases <- total_confirmed_daily[nrow(total_confirmed_daily), 2]
dat_most_recent <- dat %>%
  select(
    "county" = "County Name",
    "state" = "State",
    "most_recent" = "3/27/2020"
  ) %>%
  group_by(state) %>%
  arrange(desc(most_recent)) %>%
  mutate(
    percentage_of_total = round(most_recent / total_cases, digits = 3)
  )

dat_most_recent
```

#### Raw Data

```{r data-show-4, warning=FALSE, message=FALSE}
# Get an overview on the data
glimpse_dat <- glimpse(dat) 
glimpse_dat
```

### Basic Visualization {.tabset}

#### Daily Nation-wide Total

```{r viz-1, warning=FALSE}
ggplot(total_confirmed_daily, aes(x = date, y = total_confirmed)) + 
  geom_point() + 
  theme_light() + 
  theme(
    panel.grid.minor = element_blank(),
  ) +
  ggtitle(
    label = "Total Confirmed Cases, Nation-wide, Daily",
    subtitle = "COVID-19 GLM Model Project"
  ) +
  xlab("Date") +
  ylab("Total Confirmed Cases")
```










